Variational Monte Carlo on bosonic systems

### 2018-04-03 project_1/article/article.tex
Add review of introduction
        This includes the first part of the theory section.

\begin{abstract}
    Something very abstract and clever should go here.
    \begin{figure}[h!]
        \centering
        \includegraphics[width=0.7\textwidth]{loopholes.jpg}
    \end{figure}
\end{abstract}

@@

\section{Introduction}
    We will in this project study the Variational Monte Carlo (VMC) method, and
    use it to evaluate the ground state energy of a trapped, hard sphere Bose
    gas. \cite{dubois2001bose} \cite{nilsen2005vortices}

    By building up an increasingly intricate system we eloquently illustrate the
    strength of the VMC methods. Any Monte Carlo method draws on the fact that
    the computer will not complain when being tasked to perform a very similar
    task up to several million times in rapid succession. In general Monte Carlo
    methods is in essence a series of guesses where the current guess is only a minor
    alteration of the one prior to it. The trick lies in only accepting the subsequent
    guesses that improves upon our configuration up to a certain likelihood.

    The method we employed is called a \emph{Variational} Monte Carlo method,
    because we make an educated guess at a wavefunction which we allow to contain
    certain \emph{variational parameters}. We shall see that the system converges to a proper
    \footnote{Global, as far as we can tell.} minimum for only a certain set of parameters.

    First, we start by explaining our theoretical quantum system of a bosonic gas trapped
    in a potential well. Here we also outline the principle under which particles in the gas are
    allowed to interact. Moreover, we introduce the concept of \emph{local energy} and
    then quantum \emph{drift force}. Here we also outline in more detailed algebra the
    nature of a non-interaction system. For this system we are able to derive an exact
    variational energy to which we can compare the results from our simulations. Naturally,
    we furthermore give a description of the particle-particle interaction implemented as if each
    particle is a hard sphere. Second, we provide a thorough description of algorithms and
    statistical methods employed in this study. Here we introduce the Metropolis-Hastings
    algorithm. This algorithm can by a deft trick be expanded with the method of
    \emph{importance sampling} which make the way in which we pick the next step in
       the Monte-Carlo cycling much more intelligent. The main point of note with regards to
       the statistical analysis is that the data generated by the Monte Carlo cycling is
       autocorrelated and computation of variance must be done with the \emph{blocking method}.
       As an automated way of finding the optimal variational parameter, we sketch out the
       method of \emph{gradient descent}. Third, we put the machinery described in the
       previous sections to work. We test against the actual values in a non-interacting system
       before analysing the more interesting perturbed-trap system with interacting particles.
       Lastly, we highlight the more interesting aspects of this study in a discussion of the results
       before ending on some summary remarks.

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap, where the oscillatory frequency is different in $z$-direction so that
    $\omega_z \neq \omega$.
    The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ the \textit{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

>>>>

\begin{abstract}
    In this project we use Quantum Variational Monte Carlo on bosonic systems.
    \begin{figure}[h!]
        \centering
        \includegraphics[width=0.7\textwidth]{loopholes.jpg}
    \end{figure}
\end{abstract}

@@

\section{Introduction}
    We will in this project study the Variational Monte Carlo method, and use it
    to evaluate the ground state energy of a trapped, hard sphere Bose gas.
    \cite{dubois2001bose} \cite{nilsen2005vortices}

    By building up an increasingly intricate system we eloquently illustrate the
    strength of the Variational Monte Carlo methods. Any Monte Carlo method
    draws on the fact that the computer will not complain when being tasked to
    perform a very similar task up to several million times in rapid succession.
    Monte Carlo methods is in essence a series of guesses where the current
    guess is only a minor alteration of the one prior to it. The trick lies in
    only accepting the subsequent guesses that improves upon our configuration
    up to a certain likelihood.

    The method we have employed is called a \emph{Variational} Monte Carlo
    method, because we make an educated guess at a trial wavefunction which we
    allow certain degrees of freedom called \emph{variational parameters}. We
    shall see that the system converges to a proper \footnote{Global, as far as
    we can tell.} minimum for only a certain set of parameters.

    First, we start by explaining our theoretical quantum system of a bosonic
    gas trapped in a potential well. Here we also outline the principle under
    which particles in the gas are allowed to interact. Moreover, we introduce
    the \emph{local energy} and the quantum \emph{drift force}. Here we also
    outline in more detailed algebra the nature of a non-interacting system. For
    this system we are able to derive an exact variational energy to which we
    can compare the results from our simulations. We furthermore give a
    description of the particle-particle interaction implemented as if each
    particle is a hard sphere. Second, we provide a thorough description of
    algorithms and statistical methods employed in this study. Here we introduce
    the Metropolis-Hastings algorithm. This algorithm can by a deft trick be
    expanded with the method of \emph{importance sampling} yielding a more
    intelligent choice for the next step in the Monte Carlo steps.

    The main point of note with regards to the statistical analysis is that the
    data generated by the Monte Carlo cycling is autocorrelated and computation
    of variance must be done with the \emph{blocking method}.  As an automated
    way of finding the optimal variational parameter, we sketch out the method
    of \emph{gradient descent}.  Third, we put the machinery described in the
    previous sections to work. We test against the actual values in a
    non-interacting system before analysing the more interesting perturbed-trap
    system with interacting particles.  Lastly, we highlight the more
    interesting aspects of this study in a discussion of the results before
    ending on some summary remarks.

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        v(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    oscillator trap.  In the latter potential the oscillatory frequency is
    different in the $z$-direction, i.e., $\omega_z \neq \omega$.  The full
    Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one-body operator, $h(\vf{r}_i)$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + v(\vf{r}_i),
    \end{align}
    for equal masses, and the two-body interaction operator, $w(\vf{r}_i,
    \vf{r}_j)$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
        \label{eq:two-body_interaction}
    \end{align}
    where $a$ is the hard sphere of the particle.  The trial wavefunction,
    $\Psi_T(\vf{r})$, we will be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ is a variational parameter and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\Phi_T(\vf{r})$ is the \textit{Slater permanent} consisting of the $N$
    first single particle wavefunctions, and the correlation wavefunction, $f$,
    given by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
        \label{eq:correlation_wavefunction}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$. In practice \autoref{eq:two-body_interaction} is
    unnecessary as the situtation where $r_{jk} \leq a$ automatically yields a
    probability density of zero from \autoref{eq:correlation_wavefunction} thus
    rejecting the state completely.
